% !TEX root = ../tssp_exams.tex
% all problems are copied to stochastic_pro 2024-11-03


% it's good to follow the strategy:
% after exams copy exam here and classify int stochastic_pro

\newpage
\thispagestyle{empty}
\section{December exam}

\subsection[2025-2026b]{\hyperref[sec:sol_kr_02_2025_2026b]{2025-2026b}}
\label{sec:kr_02_2025_2026b} % \label{ссылка сюда}


Official chief mistress of King Louis XV Madame de Pompadour was born exactly 204 years earlier, 
on 29 December 1721. Her original name was Jeanne Antoinette Poisson.


\begin{enumerate}

\item Jeanne Antoinette Poisson receives letters from Diderot and d'Alembert. 
Each of them independently may be in a good mood or bad mood equiprobably.
Every person sends letters to Jeanne Antoinette Poisson according to an independent Poisson process with intensity $\lambda_{\text{good}} = 2$
or $\lambda_{\text{bad}} = 1$ [letters per week] depending on his mood.

Let $(M_t)$ be the number of letters received by Madame de Pompadour in $t$ weeks. 

\begin{enumerate}
    \item {[2 + 3]} Find $\E(M_t)$ and $\Var(M_t)$.
    \item {[2]} Is $(M_t)$ a Poisson process? If so, find its intensity. 
    \item {[3]} Let $T_1$ be the time of the first letter. Find its cumulative distribution function $F(t)$.
\end{enumerate}

\item We select a point $(X, Y)$ randomly uniformly inside the square with vertices \((0,0)\), \((4,0)\), \((4,4)\), and \((0,4)\). 
\begin{enumerate}
    \item {[2 + 3]} Find $\E(Y \mid Y - X)$ and $\Var(Y \mid Y - X)$.
    \item {[2]} Consider the $\sigma$-algebra $\cF = \sigma(Y - X)$. 
    Provide an example of two non-trivial events from $\cF$ and two events not from $\cF$.
    \item {[3]} Find $\Cov(X, Y \mid Y - X)$.
\end{enumerate}
Hint: by definition $\Cov(X, Y \mid \cF) = \E(XY \mid \cF) - \E(X\mid \cF)\E(Y\mid \cF)$.

\item Let $S_0 = 0$. 
Each hour independently of previous history the process $S_n$ increases by $1$ with probability $0.25$, decrease by $1$ with probability $0.25$ or does not change.
Consider the first moment of time $T$ when $\abs{S_{T}} = 4$.

\begin{enumerate}
    \item {[3]} If possible find $\alpha$ such that $Q_t = S_t^2 + \alpha t$ is a martingale.
    \item {[3]} If possible find $\beta$ such that $Q_t = \exp(S_t + \beta t)$ is a martingale.
    \item {[4]} Find $\E(T)$.
\end{enumerate}
% Hint: you may use Doob's theorem without checking technical details. 

\item Consider the standard Wiener process $(W_t)_{t\geq 0}$.
\begin{enumerate}
\item {[4]} Find $\Corr(W_4 - W_2, W_3 - W_1)$.
\item {[6]} Find $\E(W_3 - W_1 \mid W_4 - W_2)$ and $\Var(W_3 - W_1 \mid W_4 - W_2)$.
\end{enumerate}

%\item Consider the process $Q_t = \cos W_t$.
%\begin{enumerate}
%\item {[6]} Find $dQ_t$ and represent $Q_t$ as a sum of a constant, an Ito integral and a Riemann integral.
%\item {[2]} Is $(Q_t)$ a martingale?
%\item {[2]} Find an ordinary differential equation for $h(t) = \E(Q_t)$. You don't neet to solve this equation. 
%\end{enumerate}

\item Consider three processes, $A_t = \int_0^t u \, dW_u, \quad B_t = \int_0^t u W_u^4 \, dW_u, \quad C_t = \int_0^t u W_u^2 \, du$.
\begin{enumerate}
    \item {[2 + 2]} Find $\E(A_t)$, $\E(C_t)$.
    \item {[2 + 4]} Find $\Var(B_t)$, $\Cov(A_t, C_t)$.
\end{enumerate}

%\item Consider the process $X_t = A_t B_t$ where $A_t = \int_0^t u \, dW_u$ and $B_t = \int_0^t W_u^2 \, dW_u$.
%\begin{enumerate}
%\item {[6]} Find $dX_t$ and hence represent $X_t$ as a sum of a constant, an Ito integreal and a Riemann integral.
%\item {[2]} Is $(X_t)$ a martingale?
%\item {[2]} Find $\E(X_t)$.
%\end{enumerate}

\item Be brave!
  Consider the stochastic process $X_t = g(t) (\exp (2026 W_t) + \exp(-2026W_t))$ where $g(t)$ is some non-random function.
  \begin{enumerate}
      \item {[6]} Find $dX_t$ and hence represent $X_t$ as a sum of a constant, an Ito integreal and a Riemann integral. 
      \item {[4]} If possible find any $g(t) \neq 0$ such that $X_t$ is a martingale.
  \end{enumerate}


\end{enumerate}



\subsection[2025-2026a]{\hyperref[sec:sol_kr_02_2025_2026a]{2025-2026a}}
\label{sec:kr_02_2025_2026a} % \label{ссылка сюда}


Сегодня по народному календарю  Евстратиев день. 
Нельзя в этот день сквернословить — не то ведьмы с неба упадут прямо на голову. 
Также нельзя веник на крыльце оставлять — не то ведьмы утащат.

\begin{enumerate}
\item The Legendary November Midterm Grader grades tests according to Poisson point process with variable intensity rate.
The first 5 weeks the rate is equal to $\lambda_1 = 1$ test per week. 
Later the rate increases to $\lambda_2 = 100$ tests per week. 
Let $X_t$ be the number of tests graded after $t$ weeks. 

\begin{enumerate}
\item {[2]} Sketch the curve $v(t) = \Var(X_t)$.
\item {[4]} Using normal approximation find the probability that LNMG will check not less than $120$ works at time $t=6$.
\item {[4]} Now $t=4.5$ weeks have already passed. Find the expected time for the next test checked.
\end{enumerate}
Note: in point (b) you may use the standard normal cdf $F()$ in your answer.

%Packs of assignments arrive independently as $P_t \sim \dPois(\lambda_1)$ with intensity $\lambda_1 = 2$ packs per week. 
%Grading a single pack is a creative and randomly timed process. 
%The grading rate is also independent of arrivals and number of graded packs is $G_t \sim \dPois(\mu_1)$ with intensity $\mu_1 = 1$ pack per 3 weeks. 
%At the beginning there are 3 ungraded packs. 
%Let $X_t$ be the total number of ungraded packs at time $t \geq 0$ (including the initial packs).

%\begin{enumerate}
%    \item {[5]} Find $\E(X_t)$ and $\Var(X_t)$. 
%    What happens to $\E(X_t)$ as $t \to \infty$? 
%    Provide a practical interpretation of this result in the context of the Professor's workload.
    
%    \item {[5]} Suppose Professor Boris realizes his responsibility and suddenly increases his grading rate to $\mu_2 = 2$ packs per week. 
%    No new assignments arrive after this point ($\lambda_2 = 0$). 
%    Let $X_{13}=1$ be the number of ungraded packs. Let $T$ be the time needed to grade all the remaining packs. 
%    Derive the conditional distribution of $T$ given that $X_{13}=1$ and find $\E(T \mid X_{13}=1)$.
% \end{enumerate}

\item The joint distribution of $X$ and $Y$ is given by the table:
\begin{center}
\begin{tabular}{c|ccc}
 & $Y = -1$ & $Y = 0$ & $Y = 1$ \\
\hline
$X = -1$ & $1/8$ & $1/16$ & $1/8$ \\
$X = 0$ & $1/16$ & $1/4$ & $1/16$ \\
$X = 1$ & $1/8$ & $1/16$ & $1/8$ \\
\end{tabular}
\end{center}
Three students are building statistical models with different information sets:
\begin{itemize}
\item Roman Bokhyan: $\cF_1 = \sigma(X, Y)$, the $\sigma$-algebra generated by $X$ and $Y$. 
\item Mikhail N: $\cF_2 = \sigma(X + Y, X - Y)$.
\item Andrey L: $\cF_3 = \sigma(X^2 + Y^2)$.
\end{itemize}
\begin{enumerate}
    \item {[3]} For each pair of $\sigma$-algebras above determine whether one is contained in the other, they are equal, or neither. 
    Justify your claims.
    
    \item {[2]} Andrey and Mikhail decide to share their information. 
    Compute the number of elements in the smallest $\sigma$-algebra containing both $\cF_2$ and $\cF_3$. 
    
    \item {[5]} Help Andrey and compute $\E(X \mid \cF_3)$, $\Var(X \mid \cF_3)$.    
\end{enumerate}

\item The process $S_n$ models the «accumulated wrath of the heavens». Let $S_0 = 0$. 
Each hour independently of previous history $S_n$ increases by $1$ with probability $1/3$, decrease by $1$ with probability $1/6$ or does not change.

The witches fall from the sky when $S_n$ first reaches the level of $4$. 
Let $T = \min\{n:S_n = 4\}$.

\begin{enumerate}
    \item {[4]} For which $\alpha \neq 1$ is the process $M_n = \alpha^{S_n}$ a martingale?
    \item {[2]} Check whether $Q_n = S_n - n / 6$ is a martingale.
    \item {[4]} Find $\E(T)$.
\end{enumerate}
Hint: you may use Doob's theorem without checking technical details. 

\item Consider three processes, 
\[
A_t = \int_0^t W_u^2 \, dW_u, \quad B_t = \int_0^t W_u^4 \, dW_u, \quad C_t = \int_0^t W_u^2 \, du.
\]
\begin{enumerate}
    \item {[2 + 2]} Find $\E(A_t)$, $\E(C_t)$.
    \item {[2 + 2 + 2]} Find $\Var(A_t)$, $\Cov(A_t, B_t)$, $\Var(C_t)$.
\end{enumerate}

\item Be brave!
\begin{enumerate}
\item {[2 + 3]} Find $\E(W_3^4)$, $\E(W_3^4 \mid W_2 = 10)$.
\item {[5]} Find $\Cov(W_1 W_2, W_5 W_6)$.
\end{enumerate}

\item Consider the process
\[
R_t = \exp\left(\int_0^t u^3 dW_u + t^2\right).
\]
\begin{enumerate}
\item {[6]} Write $R_t$ as a sum of a constant, an Ito's integral and a Riemann integral.
\item {[2]} Is $(R_t)$ a martingale?
\item {[2]} Provide an ordinary (non-stochastic) differential equation for $h(t) = \E(R_t)$. 
You don't need to solve this equation. 
\end{enumerate}
Hint: you may write $R_t$ as $R_t = \exp(Q_t)$ and apply Ito's lemma :)


\end{enumerate}




\subsection[2024-2025]{\hyperref[sec:sol_kr_02_2024_2025]{2024-2025}}
\label{sec:kr_02_2024_2025} % \label{ссылка сюда}


In this text $(W_t)$ denotes the standard Wiener process\footnote{Броуновское движение [ещё пока не запрещено на территории РФ]}.


\begin{enumerate}
    \item {[10]} We keep our promises! 
    \begin{enumerate}
        \item {[3]} What is the distribution of $W_7 + 2W_8$? 
        % \item {[2]} Find the probability $\P(W_7 + 2W_8 > 1)$ in terms of a standard normal cdf $F()$.
        \item {[5]} What is the conditional distribution of $(W_7 + 2W_8 \mid W_1 = 2)$?
        \item {[2]} Find the probability $\P(W_7 + 2W_8 > 1 \mid W_1 = 2)$ in terms of a standard normal cdf $F()$.
    \end{enumerate}


    \item {[10]} Consider processes $X_t = \int_0^t W_u^3 dW_u$ and $Y_t = \int_0^t W_u^4 du$.
    \begin{enumerate}
        %\item {[2]} Find $dS_t$.
        %\item {[2]} Is $(S_t)$ a martingale? Why?
        \item {[3 + 3]} Find $\E(X_t + Y_t)$ and $\Var(X_t)$.
        \item {[4]} Find $\Cov(X_t, W_t)$.
    \end{enumerate}

    \item {[10]} Let $X_t = (W_t + g(t))\exp(-W_t - t/2)$ with $X_0 = 0$.
    \begin{enumerate}
        \item {[4 + 2]} Find $dX_t$ and write $X_t$ as a sum of two integrals.
        \item {[4]} Find at least one function $g(t)$ such that $X_t$ is a martingale.
    \end{enumerate}

    % \item Consider the Black and Scholes model with riskless rate $r$, volatility $\sigma$ and initial share price $S_0$. 

    % Let $\P^*$ denote the risk-neutral probability that makes discounted share price process a martingale.
    % \begin{enumerate}
    %     \item {[3]} Find $\P^*(S_T > 1)$.
    %     \item {[7]} Find the current price $X_0$ of an option that pays you $X_T = \max\{0, \log S_T\}$ at fixed time $T$.
    % \end{enumerate}
    
    % Hint: you may use the standard normal cumulative distribution function in your answers.
    
    \item {[10]} It's the midterm exam.
    Five students are sitting in the last row. 
    The student in the middle of the last row is the only one who brought a calculator. 
    
    Every minute his calculator moves one seat to the right (+1) or one seat to the left (-1) with equal probabilities.

    Let's $X_n$ be the coordinate of the calculator at time $n$ with $X_0 = 0$.
    \begin{enumerate}
        \item {[4]} Check whether $M_n = X_n^2 - n$ is a martingale.
        \item {[6]} Find the average time for the calculator to reach the end of the row (left or right).
    \end{enumerate}

    \item {[10]} Consider a two-period binomial tree model with an initial share price $S_0 = 100$. 
    The up and down share price multipliers are $u = 2$ and $d = 0.5$. 
    Risk-free interest rate is $r = 10\%$ in the first period. 
    The central bank will increase the interest rate in the second period exactly to $r = 20\%$.

    The option is a European call with strike price $K = 300$ and maturity $T = 2$. 
    %But at the end of the second year, all payments to the option holder will be blocked due to the economical sanctions with probability equal to 0.5.
    
    \begin{enumerate}
        \item {[4]} Find the risk neutral probabilities.
        \item {[6]} Find the arbitrage free price $X_0$ of this option.
    \end{enumerate}

    \item {[10]} The assistant has not checked the home assignments in time. 
    Hence teachers are receiving student complaints according to a Poisson process with daily rate $\lambda = 3$ messages per hour from $7:00$ to $23:00$ and 
    nightly rate $\lambda = 1$ message per hour from $23:00$ to $7:00$.

    Let $X$ be the number of messages from $6:00$ to $9:00$ and $Y$ — number of messages from $7:00$ to $9:00$.
    \begin{enumerate}
        \item {[2+2+2]} Find $\E(X)$, $\Var(X)$ and $\P(X = 2)$.
        \item {[4]} Find the conditional distribution of $Y$ given that $X = 2$.
    \end{enumerate}
% \item мартингал в непрерывном времени? дуб?


\end{enumerate}



\subsection[2023-2024]{\hyperref[sec:sol_kr_02_2023_2024]{2023-2024}}
\label{sec:kr_02_2023_2024} % \label{ссылка сюда}

\begin{enumerate}
    \item {[10 points]} Let $(X_t)$ be independent identically distributed random variables 
    with $\P(X_i = -1) = 0.4$ and $\P(X_i = +1) = 0.6$. Consider the sum $S_t = X_1 + X_2 + \ldots + X_t$.
 \begin{enumerate}
    \item {[3]} Is $S_t$ a martingale?
    \item {[7]} Find all constants $c$ such that $M_t = c^{S_t}$ is a martingale. 
 \end{enumerate}

\item {[10 points]} Let $a(t)$ be a deterministic function, $M_t = a(t) \cos (3W_t)$ and $(W_t)$ is a Wiener process.

\begin{enumerate}
    \item {[4]} Find $dM_t$. 
    \item {[6]} Find a non-zero function $a(t)$ such that $(M_t)$ is a martingale.
\end{enumerate}
 
\item {[10 points]} You have two correlated Wiener processes, $(A_t)$ and $(B_t)$, with $\Corr(A_t - A_s, B_t - B_s) = \rho$ for all $t > s$.

Split the time interval $[0;t]$ into $n$ small segments of equal length. 
Let $\Delta^A_i$ be the increment of the Wiener process $(A_t)$ on the $i$-th small segment, i.e. $\Delta^A_i = A(it/n) - A((i-1)t/n)$.
Let $\Delta^B_i$ be the increment of the Wiener process $(B_t)$ on the $i$-th small segment.

Consider the sum of cross-products, $S_n = \sum_{i=1}^n \Delta^A_i \Delta^B_i$.
\begin{enumerate}
    \item {[3]} Find $\E(S_n)$.
    \item {[4]} Does $\Var(S_n)$ tend to $0$ when $n \to \infty$?
    \item {[2]} Find the mean square limit of $S_n$. 
    \item {[1]} How would you write this limit in a short hand notation with $dA_t$ and $dB_t$?
\end{enumerate}

\item {[10 points]} The process $(X_t)$ has $X_0 = 2024$, $dX_t = W_t^2 dW_t + W_t dt$, where $(W_t)$ is a Wiener process.
\begin{enumerate}
    \item {[2]} Is $(X_t)$ a martingale?
    \item {[4]} Find $d(X_t W_t)$.
    \item {[4]} Find $\Cov(X_t, W_t)$.
\end{enumerate}

\item {[10 points]} Consider two-period binomial model with initial share price $S_0 = 600$. 
Up and down share price multipliers are $u=1.2$, $d=0.9$, risk-free interest rate is $r = 0.05$ per period. 

The option pays you the maximal share price $X_2 = \max\{S_0, S_1, S_2\}$ at $t=2$.

\begin{enumerate}
    \item {[4]} Find the risk neutral probabilities. 
    \item {[6]} Find the current price $X_0$ of this option. 
%    \item How many shares should I have at $t=1$ in the «down» state of the world to replicate the option?
\end{enumerate}

\item {[10 points]}  Consider the framework of Black and Scholes model with riskless rate $r$, volatility $\sigma$ and initial share price $S_0$. 

Find the current price $X_0$ of an option that pays you $X_T = 1$ at fixed time $T$ if $S_T \geq 2 S_0$.

Hint: you may use the standard normal cumulative distribution function in your answer.


\end{enumerate}



\subsection[2022-2023]{\hyperref[sec:sol_kr_02_2022_2023]{2022-2023}}
\label{sec:kr_02_2022_2023} % \label{ссылка сюда}



Short rules: 120 minutes, you may use two A4 cheat-sheets, offline + one online group.

\begin{enumerate}

\item Consider $X_t = \int_0^t W_u^3 dW_u + \int_0^t (W_u^3 + 3W_u u ) du - W_t^3 \cdot t$.

\begin{enumerate}
    \item Find $dX_t$ and the corresponding full form. 
    \item Is $X_t$ a martingale?
\end{enumerate}

\item Consider $X_t = \exp(-2W_t - 2t)$.
\begin{enumerate}
    \item Find $dX_t$. Is $X_t$ a martingale?
    \item Find $\E(X_t)$ and $\Var(X_t)$.
    \item Find $\int_0^t X_u dW_u$.
\end{enumerate}

\item As usual $(W_t)$ is a Wiener process.
\begin{enumerate}
    \item Find $\E(W_5 W_4 \mid W_4)$, $\Var(W_5 W_4 \mid W_4)$.
    \item Find covariance $\Cov(W_4 W_5, W_5 W_6)$.
\end{enumerate}

\item Let $X_i$ be independent identically distributed with $\P(X_i = 1) = 0.9$, $\P(X_i = -1 ) = 0.1$. 

Find all constants $a$ and $b$ such that $Y_t = a \exp\left(b\sum_{i=1}^t X_i\right)$ is a martingale. 

\item Consider two-period binomial model with initial share price $S_0 = 600$, 
Up and down multipliers are $u=1.2$, $d=0.9$, risk-free interest rate is $r = 0.05$ per period. 

Consider an option that pays you $X_2 = 100$ at $T=2$ if $S_2 > S_1$ and nothing otherwise. 

\begin{enumerate}
    \item Find the risk neutral probabilities. 
    \item Find the current price $X_0$ of the asset. 
    \item How much shares should I have at $t=1$ in the «up» state of the world to replicate the option?
\end{enumerate}

\item Consider Black and Scholes model with riskless rate $r$, volatility $\sigma$ and initial share price $S_0$. 

Find the current price $X_0$ of an option that pays you $X_2 = S_1^3$ at time $T=2$. 

\end{enumerate}




\subsection[2021-2022]{\hyperref[sec:sol_kr_02_2021_2022]{2021-2022}}
\label{sec:kr_02_2021_2022} % \label{ссылка сюда}

Short rules: 120 minutes, online without proctoring, $(W_t)$ is a standard Wiener process.

Date: 2021-12-25

\begin{enumerate}

  \item (10 points) Consider an Ito's process $I_t = 2022 + W_t t^2 + \int_0^t W_u^3 dW_u + \int_0^t W_u^2 du$.
    \begin{enumerate}
      \item Find $dI_t$ and check whether $I_t$ is a martingale. 
      \item Check whether $J_t = I_t - \E(I_t)$ is a martingale.
    \end{enumerate}
    
  \item (10 points) The random variables $(Z_t)$ are independent identically distributed 
  with moment generating function given by $M_{Z}(u) = 1/(1 - 5u)^3$. 
  
  We define $X_t$ as $X_t = \exp(Z_1 + 2Z_2 + 3Z_3 + \ldots + tZ_t)$ with $X_0 = 0$. 
  
  If possible find a martingale of the form $Y_t = h(t) X_t$ where $h()$ is a non-random function.
  
  \item (10 points) The process $(Z_t)$ in discrete time is called \textit{stationary} if it has constant expected value 
  and constant covariances $\gamma_k$ that do not depend on $t$. 
  \[
  \begin{cases}
  \E(Z_t) = \mu; \\
  \Cov(Z_t, Z_t) = \gamma_0; \\
  \Cov(Z_t, Z_{t+1}) = \gamma_1; \\
  \Cov(Z_t, Z_{t+2}) = \gamma_2; \\
  \ldots \\
  \end{cases}
  \]
  
  \begin{enumerate}
    \item If possible provide an example of a martingale that is not stationary.
    \item If possible provide an example of a stationary process that is not a martingale.
  \end{enumerate}
  
  \item (10 points) Find $\E(W_1 W_2 W_3)$ and $\E(W_2 W_3 \mid W_1)$.
  \item (10 points) Ded Moroz would like to receive $X_T = S^{-1}_T$ at time $T$ if $S_T < 1$ and nothing otherwise.
      
  Assume the framework of Black and Scholes model, $S_t$ is the share price, $r$ is the risk free rate,
  $\sigma$ is the volatility. 
  
  How much Ded Moroz should pay now at $t=0$?
  
  
  \item (20 points) Martingales are everywhere :)
  
  Consider the process $Y_t = \exp( - u W_t)$. 
  
  \begin{enumerate}
    \item Find a multiplier $h(u, t)$ such that $M_t = h(u, t) \cdot Y_t$ is a martingale. 
    \item Find $dY_t$, $\E(Y_t)$ and $\Var(Y_t)$.
    \item Consider $M_t$ that you have found as a function of $u$. 
    Find the Taylor approximation of the function $M_t(u)$ up to $u^4$. 
    \item Consider the coefficient before $u^4$ in the Taylor expansion of $M_t(u)$. 
    Is it a martingale?
  \end{enumerate}
  
  \item Bonus point. Guess your exam result (out of 70 possible points). 
  
  \end{enumerate}
  
      

% \subsection[что идет в оглавление]{\hyperref[на что ссылка]{текст ссылки}}
\subsection[2020-2021]{\hyperref[sec:sol_kr_02_2020_2021]{2020-2021}}
\label{sec:kr_02_2020_2021} % \label{ссылка сюда}



Today we celebrate Christmas Eve and 78 years of the Narkompros (People's Commissariat for Education) order governing the compulsory use of the letter «ё» in education process.

Date: 2020-12-24

\begin{enumerate}

    \item Ded Moroz would like to receive $S_1^3$ roubles at time $T=2$,
    where $S_t$ is the share price. Assume Black-Schёles model is valid, 
    the risk-free rate is $r=0.1$ and current share price is $S_0=100$.

    How much Ded Moroz should pay now at $t=0$?
    
    \item Consider stationary $AR(2)$ model, $y_t = 2 + 0.3 y_{t-1} - 0.02 y_{t-2} + u_t$, where $(u_t)$ is a white noise
    with $\Var(u_t) = 4$.
    
    The last two observations are $y_{100} = 2$, $y_{99} = 1$.
    \begin{enumerate}
        \item Find 95\% predictive interval for $y_{102}$.
        \item Find the first two values of the autocorrelation function, $\rho_1$, $\rho_2$.
        \item Find the first two values of the partial autocorrelation function, $\phi_{11}$, $\phi_{22}$.
    \end{enumerate}

    Hint: you need no more than 10 seconds to find both partial autocorrelations provided (b) is sёlved.

    \item The process $y_t$ is described by a simple $GARCH(1, 1)$ model:
    \[ 
        \begin{cases}
            y_t = \sigma_t \nu_t \\
            \sigma_{t}^{2}= 1 + 0.2 y_{t-1}^{2}+ 0.3 \sigma_{t-1}^{2}    \\
            \nu_t \sim \cN(0;1)
        \end{cases}     
    \]

    The variables $\nu_t$ are independent of past variables $y_{t-k}$, $\nu_{t-k}$, $\sigma_{t-k}$ for all $k\geq 1$.
    The prёcesses $y_t$, $\sigma^2_t$ are stationary. 


    Given $\sigma_{100}=1$ and $\nu_{100} = 0.5$ find 95\% predictive interval for $y_{102}$. 


    \item Snegurochka studies a stochastic analog of the Fibonacci sequence
    \[
        y_t = y_{t-1} + y_{t-2} + u_t,
    \]
    where $(u_t)$ is a white noise process. 
    \begin{enumerate}
        \item How many non-stationary solutions are there?
        \item What can you say about the number and the structure of the stationary solutions?
        \item Can Snёgurochka find two starting constants $y_0 = c_0$ and $y_1=c_1$ in such a way to make a solution stationary?
    \end{enumerate}

    Be brave! There are two more exercises!

    \item The semi-annual $y_t$ is modelled by $ETS(AAA)$ process:
    
    \[
    \begin{cases}
        u_t \sim \cN(0; 4) \\
        s_t = s_{t-2} + 0.1 u_t \\
        b_t = b_{t-1} + 0.2 u_t \\
        \ell_t = \ell_{t-1} + b_{t-1} + 0.3 u_t \\
        y_t = \ell_{t-1} + b_{t-1} + s_{t-2} + u_t \\
    \end{cases}    
    \]

    \begin{enumerate}
        \item Given that $s_{100} = 2$, $s_{99} = -1.9$, $b_{100} = 0.5$, $\ell_{100} = 4$ find 95\% prёdictive interval for $y_{102}$. 
        \item In this problem particular values of parameters are specified. And how many parameters are estimated in semi-annual $ETS(AAA)$ model before real forecasting?
    \end{enumerate}

    \item The variables $x_t$ take values $0$ or $1$ with equal probabilities.
    The variables $u_t$ are normal $\cN(0; 1)$. All variables are independent.
    
    Consider the process  $z_t = x_t (1-x_{t-2}) u_t$.

    \begin{enumerate}
        \item Find the covariance $\Cov(z_t, z_s)$. Is the process $z_t$ stationary?
        \item Given that $z_{100} = 2.3$ find shёrtest predictive intervals for $z_{101}$ and $z_{102}$ with probability of coverage at least 95\%.
    \end{enumerate}

    \item[Bёnus:] How many letters «ё» have you spotted?
    
 
\end{enumerate}
